Follow-up Samples

### 2017-02-23 submission/manuscript_outline_20161024.tex

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with previous research suggesting that a link may exist
between the fecal bacterial microbiome and CRC. The overall objective of
our study was to test the hypothesis that changes in the bacterial
microbiome occur after lesion (i.e.~adenoma or carcinoma) removal.
Specifically, we wanted to identify what within the community was
different before and after removal of said lesion.

\textbf{Results:} The bacterial microbiome changed more in response to
lesion removal in carcinoma cases compared to adenoma cases (P-value
\textless{} 0.05). There was no difference for either the adenoma or
carcinoma group in the relative abundance of any OTU between pre and
post lesion removal (P-value \textgreater{} 0.05). A model built to
classify lesion had an AUC range of 0.811 - 0.866 while a model built to
classify initial samples had an AUC range of 0.641 - 0.805. Follow up
samples had a decrease in the positive probability of lesion (P-value
\textless{} 0.05) suggesting a movement towards a more normal bacterial
community. The initial sample model had a decrease in positive
probability for the follow up samples to be an initial sample (P-value
\textless{} 0.05). The lesion model used a total of 53 variables while
the initial sample model used a total of 70 variables. A total of 23
OTUs were common to both models with the majority of these classifying
to commensal bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales},
\emph{Blautia}, and \emph{Ruminococcaceae}).

@@

Over the last few years a few studies have shown how this bacterial
microbiome {[}3{]} or specific members within it {[}4{]} could be
directly involved with the pathogenesis of CRC. There has also been
research into how the bacterial microbiome could be altered directly on
tumor tissue itself {[}5{]}. These studies have helped to provide a
tantalizing link between the bacterial microbiome and CRC. Building off
of these findings, there has been promising work on the bacterial
microbiome and its ability to complement existing screening methods such
as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
{[}6,7{]}. Although these studies suggest that the bacterial microbiome
might change after treatment there remains limited information on the
bacterial microbiome response to lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) removal of lesion
(adenoma or carcinoma). We analyzed changes in alpha and beta diversity
as well as specific Operational Taxonomic Units (OTUs) relative
abundance. We utilized Random Forest to build two models. The first
model was built to classify lesion and normal samples while the second
was built to classify initial and follow up samples. Subsequent
observations on how these models, as well as specific OTUs within them,
performed pre and post lesion removal helped inform us as to whether
initial and follow up samples were changing and whether it was towards a
more normal bacterial microbiome or not. We also investigated the two
models for similar important OTUs to identify the crucial OTUs for not
only classifying lesion or normal samples but also initial and follow
up.

@@

\textbf{\emph{Carcinoma Associated Bacteria:}} Previous literature has
suggested that a number of oral microbes may be important in CRC
pathogenesis {[}8{]}. So we next examined whether there were changes in
previously well described carcinoma associated OTUs, such as
\emph{Porphyromonas asaccharolytica} (Otu000202), \emph{Fusobacterium
nucleatum} (Otu000442), \emph{Parvimonas micra} (Otu001273), and
\emph{Peptostreptococcus stomatis} (Otu001682). We first observed that
only a small percentage of those with adenoma or carcinoma were positive
or had an appreciable relative abundance for any of these respective
OTUs {[}Figure 2{]}. Despite this, those with carcinoma had a decrease
in relative abundance from initial to follow up for \emph{Parvimonas
micra} (P-value \textless{} 0.05) and \emph{Porphyromonas
asaccharolytica} (P-value \textless{} 0.05) {[}Figure 2A{]}. In
contrast, there was no difference in relative abundance in any of these
OTUs when considering only those with adenoma {[}Figure 2B{]}.

\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
were any common bacterial microbiome changes in individuals with adenoma
or carcinoma versus normal controls. We investigated this by creating a
model to classify samples as lesion versus normal based on the bacterial
community and FIT measurements. This model had an AUC range of 0.811 -
0.866 after 100 iterations of 20 repeated 10-fold cross validations. The
ROC curve for the final lesion model used was within the observed range
of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
a total of 53 variables that were used in this model {[}Figure 3B{]}.
The FIT measurement for fecal blood resulted in the largest decrease in
MDA while the OTU with the largest MDA was \emph{Lachnospiraceae}
(Otu000015) {[}Figure 3B{]}.

@@

\textbf{\emph{The Initial Sample Model:}} After building a model to
classify based on lesion we built a separate model specifically to be
able to classify whether samples were initial samples based on the
bacterial community and FIT measurements. The initial sample model had
an AUC range of 0.641 to 0.805 after 100 iterations of 20 repeated
10-fold cross validations. The test set AUC range for this model
performed better then the training set AUCs. There was a marked decrease
in the ROC curve for the final model used when compared to the 100 test
set AUC iterations {[}Figure 4A{]}. There were a total of 70 variables
that were used for this model {[}Figure 4B{]}. The variable that
resulted in the largest MDA was \emph{Pseudomonas} (Otu000438) while FIT
measurement for fecal blood resulted in the sixth largest decrease in
MDA {[}Figure 4B{]}.

@@

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool
{[}6,7,9--11{]} by exploring what happens to the bacterial community
after surgical removal of a lesion. Based on previous work by Arthur, et
al. {[}12{]} it may not be surprising to have E.coli as one of the most
important OTUs and one that was common to both models. Interestingly,
many of the most important OTUs had taxonomic identification for
resident gut microbes. This could suggest that the bacterial community
is one of the first components that could change during the pathogenesis
of disease. These bacterial microbiome changes could be the first step
in allowing more inflammatory bacterium to gain a foothold within the
colon {[}8{]}.

@@

We next created a model that incorporated FIT and the bacterial
microbiome to be able to classify lesions (adenoma or carcinoma). Based
on this model we found that the follow up samples were closer to normal
then the initial samples due to a decrease in positive probability for
lesion and that the commonly associated CRC bacteria were not highly
represented within this model with the exception of \emph{Porphyromonas
asaccharolytica} {[}Figure 3{]}. Although there was a detectable change
and this change was towards what would be expected for normal controls,
it should be noted that the follow up samples may not be a completely
normal bacterial microbiome.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after surgery {[}Figure 3C \&
3D{]}. It should be noted that our study does not argue against the
importance of these CRC associated bacteria in the pathogenesis of
disease but rather that they are not the main bacteria changing after
surgical removal of lesion. In fact, it is possible that these CRC
associated bacteria are important in the transition from adenoma to
carcinoma and would be one explanation as to why in our data we not only
see high initial relative abundances in carcinoma and not adenoma
individuals but also large decreases in relative abundance in some of
those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

@@

\textbf{\emph{Study Design and Patient Sampling:}} The sampling and
design were similar to that reported in Baxter, et al {[}6{]}. In brief,
study exclusion involved those who had already undergone surgery,
radiation, or chemotherapy, had colorectal cancer before a baseline
fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
microbiome has been shown to normalize within this time period {[}18{]}.
Our follow up data set had a total of 67 individuals that not only had a
sample as described but also a follow up sample between 188 - 546 days
after lesion removal and treatment had been completed. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

@@

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}21{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main training and testing of the model
completed on a data set of 423 individuals. This model was then applied
to our follow up data set of 67 individuals.

@@

\textbf{\emph{Initial Sample Model Creation:}} We also investigated
whether a model could be created that could identify pre (initial) and
post (follow up) lesion removal samples from each other. The main
difference was that only the 67-person cohort was used at all stages of
model building and classification. Other than this difference the
creation of this model and optimization of the mtry hyper-parameter was
completed using the same procedure as was used for the lesion model.
Instead of classifying samples as positive or negative of lesion this
model classified samples as positive or negative for being an initial to
lesion removal sample.

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}22{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate a Benjamini-Hochberg (BH) correction was applied {[}23{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected.

\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
any differences based on whether the individual had an adenoma or
carcinoma. This was done by testing initial and follow up samples for
differences in alpha and beta diversity, testing differences in FIT
between initial and follow ups, testing all OTUs, and investigating the
relative abundance of specific previously associated CRC bacteria
(\emph{Fusobacterium nucleatum}, \emph{Parvimonas micra},
\emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
stomatis}) based on adenoma and carcinoma. From here the lesion model
was then tested for accuracy in prediction and whether it reduced the
positive probability of lesion after surgery. We then used the initial
sample model to assess whether it could classify samples better then the
lesion model and whether it could reduce the positive probability of an
initial sample in the follow up samples. Finally, a list of common OTUs
were found for the two two different models used.

@@

\textbf{Figure 1: General Differences between the Adenoma and Carcinoma
Group.} A) A significant difference was found between the adenoma and
carcinoma group for thetayc (P-value = 0.000472). B) A significant
difference was found between the adenoma and carcinoma group for change
in FIT measurement (P-value = 2.15e-05). C) NMDS of the initial and
follow up samples for the Adenoma group. D) NMDS of the initial and
follow up samples for the Carcinoma group. For C) and D) the teal
represents initial samples and the pink represents follow up samples.

>>>>

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with previous research suggesting that a link may exist
between the fecal bacterial microbiome and CRC. The overall objective of
our study was to test the hypothesis that changes in the bacterial
microbiome occur after lesion (i.e.~adenoma or carcinoma) removal.
Specifically, we wanted to identify components within the community that
were different before and after removal of said lesion.

\textbf{Results:} The bacterial microbiome changed more in response to
lesion removal in carcinoma cases compared to adenoma cases (P-value
\textless{} 0.05). There was no difference for either the adenoma or
carcinoma group in the relative abundance of any OTU between pre and
post lesion removal (P-value \textgreater{} 0.05). A model built to
classify lesion had an AUC range of 0.811 - 0.866 and follow up samples
had a decrease in the positive probability of lesion (P-value
\textless{} 0.05) suggesting a movement towards a more normal bacterial
community. A second model built to classify initial samples had an AUC
range of 0.641 - 0.805 and had a decrease in positive probability for
the follow up samples to be an initial sample (P-value \textless{}
0.05). The lesion model used a total of 53 variables while the initial
sample model used a total of 70 variables. A total of 23 OTUs were
common to both models with the majority of these classifying to
commensal bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales},
\emph{Blautia}, and \emph{Ruminococcaceae}).

@@

Over the last few years studies have shown how the bacterial microbiome
{[}3{]} or specific members within it {[}4{]} could be directly involved
with the pathogenesis of CRC. There has also been research into how the
bacterial microbiome could be altered directly on tumor tissue itself
{[}5{]}. These studies have helped to provide a tantalizing link between
the bacterial microbiome and CRC. Building off of these findings, there
has been promising work on the bacterial microbiome and its ability to
complement existing screening methods such as Fecal Immunoglobulin Test
(FIT) or act alone as a screening tool {[}6,7{]}. Although these studies
suggest that the bacterial microbiome might change after treatment there
remains limited information on the bacterial microbiome response to
lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) removal of lesion
(adenoma or carcinoma). We analyzed changes in alpha and beta diversity
as well as the relative abundance of specific Operational Taxonomic
Units (OTUs). We utilized Random Forest to build two models. The first
model was built to classify lesion and normal (non-lesion) samples while
the second was built to classify initial and follow up samples.
Subsequent observations on how these models, as well as specific OTUs
within them, performed pre and post lesion removal helped inform us as
to whether initial and follow up samples were changing and whether it
was towards a more normal bacterial microbiome or not. We also
investigated the two models for similar important OTUs to identify the
crucial OTUs for not only classifying lesion or normal samples but also
initial and follow up.

@@

\textbf{\emph{Carcinoma Associated Bacteria:}} Previous literature has
suggested that a number of oral microbes may be important in CRC
pathogenesis {[}8{]}. So we next examined whether there were changes in
previously well described carcinoma associated OTUs, such as
\emph{Porphyromonas asaccharolytica} (Otu000202), \emph{Fusobacterium
nucleatum} (Otu000442), \emph{Parvimonas micra} (Otu001273), and
\emph{Peptostreptococcus stomatis} (Otu001682). We first observed that
only a small percentage of those with adenoma or carcinoma were positive
or had a relative abundance above 0.1\% for any of these respective OTUs
{[}Figure 2{]}. Despite this, those with carcinoma had a decrease in
relative abundance from initial to follow up for \emph{Parvimonas micra}
(P-value \textless{} 0.05) and \emph{Porphyromonas asaccharolytica}
(P-value \textless{} 0.05) {[}Figure 2A{]}. In contrast, there was no
difference in relative abundance in any of these OTUs when considering
only those with adenoma {[}Figure 2B{]}.

\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
were any common bacterial microbiome changes in individuals with adenoma
and carcinoma versus normal controls. We investigated this by creating a
model to classify samples as lesion versus normal based on the bacterial
community and FIT measurements. This model had an AUC range of 0.811 -
0.866 after 100 iterations of 20 repeated 10-fold cross validations. The
ROC curve for the final lesion model used was within the observed range
of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
a total of 53 variables that were used in this model {[}Figure 3B{]}.
The FIT measurement for fecal blood resulted in the largest decrease in
MDA while the OTU with the largest MDA was \emph{Lachnospiraceae}
(Otu000015) {[}Figure 3B{]}.

@@

\textbf{\emph{The Initial Sample Model:}} After building a model to
classify based on lesion we built a separate model specifically to be
able to classify whether samples were initial (before lesion was
removed) samples based on the bacterial community and FIT measurements.
The initial sample model had an AUC range of 0.641 to 0.805 after 100
iterations of 20 repeated 10-fold cross validations. The test set AUC
range for this model performed better then the training set AUCs. There
was a marked decrease in the ROC curve for the final model used when
compared to the 100 test set AUC iterations {[}Figure 4A{]}. There were
a total of 70 variables that were used for this model {[}Figure 4B{]}.
The variable that resulted in the largest MDA was \emph{Pseudomonas}
(Otu000438) while FIT measurement for fecal blood resulted in the sixth
largest decrease in MDA {[}Figure 4B{]}.

@@

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool
{[}6,7,9--11{]} by exploring what happens to the bacterial community
after lesion removal. Based on previous work by Arthur, et al. {[}12{]}
it may not be surprising to have E.coli as one of the most important
OTUs and one that was common to both models. Interestingly, many of the
most important OTUs had taxonomic identification for resident gut
microbes. This could suggest that the bacterial community is one of the
first components that could change during the pathogenesis of disease.
These bacterial microbiome changes could be the first step in allowing
more inflammatory bacterium to gain a foothold within the colon {[}8{]}.

@@

We next created a model that incorporated FIT and the bacterial
microbiome to be able to classify lesions (adenoma and carcinoma). Based
on this model we found that the follow up samples were closer to normal
then the initial samples due to a decrease in positive probability for
lesion and that the commonly associated CRC bacteria were not highly
represented within this model with the exception of \emph{Porphyromonas
asaccharolytica} {[}Figure 3{]}. Although there was a detectable change
towards what would be expected for normal controls, it should be noted
that the follow up samples may not be a completely normal bacterial
microbiome.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after surgery {[}Figure 3C \&
3D{]}. It should be noted that our study does not argue against the
importance of these CRC associated bacteria in the pathogenesis of
disease but rather that they are not the main bacteria changing after
removal of lesion. In fact, it is possible that these CRC associated
bacteria are important in the transition from adenoma to carcinoma and
would be one explanation as to why in our data we not only see high
initial relative abundances in carcinoma and not adenoma individuals but
also large decreases in relative abundance in some of those with
carcinoma but not in those with adenoma after surgery {[}Figure 2{]}.

@@

\textbf{\emph{Study Design and Patient Sampling:}} The sampling and
design were similar to that reported in Baxter, et al {[}6{]}. In brief,
study exclusion involved those who had already undergone surgery,
radiation, or chemotherapy, had colorectal cancer before a baseline
fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
microbiome has been shown to normalize back to a pre-colonscopy
community within this time period {[}18{]}. Our follow up data set had a
total of 67 individuals that not only had a sample as described but also
a follow up sample between 188 - 546 days after lesion removal and
treatment had been completed. This study was approved by the University
of Michigan Institutional Review Board. All study participants provided
informed consent and the study itself conformed to the guidelines set
out by the Helsinki Declaration.

@@

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}21{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main training and testing of the model
completed on an independent data set of 423 individuals. This model was
then applied to our follow up data set of 67 individuals. It should be
noted that all individuals with an adenoma or carcinoma were grouped
together to form the lesion group and the model was not created to find
differences between normal, adenoma, and carcinoma but rather
differences between both adenoma and carcinoma versus normal.

@@

\textbf{\emph{Initial Sample Model Creation:}} We also investigated
whether a model could be created that could identify pre (initial) and
post (follow up) lesion removal samples from each other. The main
difference was that only the 67-person cohort was used at all stages of
model building and classification. Other than this difference the
creation of this model and optimization of the mtry hyper-parameter was
completed using the same procedure as was used for the lesion model.
Instead of classifying samples as positive or negative of lesion this
model classified samples as positive or negative for being an initial
sample prior to lesion removal.

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}22{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}23{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected.

\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
any differences based on whether the individual had an adenoma or
carcinoma. This was done by testing initial and follow up samples for
differences in alpha and beta diversity, testing differences in FIT
between initial and follow ups, testing all OTUs, and investigating the
relative abundance of specific previously associated CRC bacteria
(\emph{Fusobacterium nucleatum}, \emph{Parvimonas micra},
\emph{Peptostreptococcus assacharolytica}, and \emph{Porphyromonas
stomatis}) based on adenoma and carcinoma. From here the lesion model
was then tested for accuracy in prediction and whether it reduced the
positive probability of lesion after surgery. We then used the initial
sample model to assess whether it could classify samples better then the
lesion model and whether it could reduce the positive probability of an
initial sample in the follow up samples. Finally, a list of common OTUs
were found for the two different models used.

@@

\textbf{Figure 1: General Differences between the Adenoma and Carcinoma
Group.} A) A significant difference was found between the adenoma and
carcinoma group for thetayc (P-value = 0.000472). Advanced adenomas are
denoted as Screen Relevant Neoplasia (SRN). B) A significant difference
was found between the adenoma and carcinoma group for change in FIT
measurement (P-value = 2.15e-05). Advanced adenomas are denoted as
Screen Relevant Neoplasia (SRN). C) NMDS of the initial and follow up
samples for the adenoma group. D) NMDS of the initial and follow up
samples for the carcinoma group.
